---
title: "Assignment Report"
author: "Richie"
date: "5/17/2021"
output: html_document
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Data Report Markdown

Group 5

- Teo Richie (17207269/1)
- Yap Gay Chin (17204377/1)
- Jovi Koh Wei Chiang (17202879/1)
- Seh Chia Shin (17202838/1)

### Introduction 

In 2020 a significant threat to public health emerged. The novel severe acute respiratory syndrome coronavirus 2 (SARS-CoV-2) epidemic outbreak emerged in December 2019 from Wuhan City, Hubei Province, China and spread to the rest of the world. This disease was named COVID-19 by World Health Organization. To date (17th April 2020) a total of 2,230,439 cases of COVID-19; 150,810 cases of deaths and 564,210 recovered cases have been reported worldwide. In this review the SARS-CoV-2 morphology, pathogenic mechanism, similarities and differences between SARS-CoV and Middle East Respiratory Syndrome and severe acute respiratory syndrome, transmission mode, diagnosis, treatment, and preventive measures were investigated. The outbreak of COVID-19 from a Malaysian perspective was explored and mental health care during the COVID-19 outbreak was explored. To date, there is no vaccine or no specific treatment for COVID-19. Therefore, preventive measures are very important to prevent and control the rapid spread of the SARS-CoV-2 virus. Preparedness should be a priority for future pandemic outbreaks.

### Problem Statement

How many people are being diagnosed as positive, died, or recovered from COVID-19 per day?,How did the confirmed, deceased and recovered cases evolve?, What are the specific cases by country per day?", How many countries were affected by COVID-19?

### Data Acquisition

For this assignment, we will be implementing all the knowledge that we have learned to produce valuable insights using R, from data acquisition to building a simple web app using Shiny.

The data set we choose to be used for our question is a data set from John Hopkins University, it is updated daily and covers the whole world.

The data set has a a very huge size and up until now we cannot conclude the size of the data set since it is updated daily dynamically by the John Hopkins University, however, it contains 4 variables such as the **Province/State**, **Country/Region**, **Latitude**, **Longitude** and so on. 

```{r}
setwd("C:/Users/Coding/Desktop/IDS Assignment/COVID-19-Dashboard")
```

In our case, there is no need for data cleaning process since the data itself was constantly checked by the John Hopkins University's data team and they ensured that the rows were thoroughly cleaned before archiving and updating the data set. Hence, all that's left to do is Data Wrangling and Data Extraction.

### Data Extraction, Wrangling and Analysis

Before we begin, we identified the dependencies that we needed for this project and imported all the relevant libraries

```{r}
library("shiny")
library("shinydashboard")
library("tidyverse")
library("leaflet")
library("plotly")
library("DT")
library("fs")
library("wbstats")

source("utils.R", local = T)
```

We firstly extract the data by downloading it from John Hopkins University GitHub page and by implementing **download.file()** function, we extract the data and paste it to the path where we have three csv files named **time_series_covid19_confirmed_global**, **time_series_covid19_deaths_global**, **time_series_covid19_recovered_global** respectively under the **data** folder.

```{r}
downloadGithubData <- function() {
  download.file(
    url      = "https://github.com/CSSEGISandData/COVID-19/archive/master.zip",
    destfile = "data/covid19_data.zip"
  )
  
  data_path <- "COVID-19-master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_"
  unzip(
    zipfile   = "data/covid19_data.zip",  
    files     = paste0(data_path, c("confirmed_global.csv", "deaths_global.csv", "recovered_global.csv")),
    exdir     = "data",
    junkpaths = T
  )
}
```

Then to ensure that the data is the latest one posted by John Hopkins University, we implemented a function that will download the data periodically to update the current data set in the directory. The function also cross checks whether the current work directory has the folder named **data** in it, if it doesn't it will create the folder. The function also implements a timer of 30 minutes interval. Every 30 minutes, it will invoke the **downloadGitHubData()** function to update the current data set.

```{r}
updateData <- function() {
  if (!dir_exists("data")) {
    dir.create('data')
    downloadGithubData()
  } else if ((!file.exists("data/covid19_data.zip")) || (as.double(Sys.time() - file_info("data/covid19_data.zip")$change_time, units = "hours") > 0.5)) {
    downloadGithubData()
  }
}
```

After that, we created 2 variables which are **data_confirmed** and **data_deceased** that reads the two corresponding csv files respectively under the data folder. To get the latest data, we used the **as.Date()** function to convert between character representations and objects of class **Date** representing calendar dates corresponding to the time-series data provided in the data set csv. After that, we created another variable **changed_date** that changes the metadata of the data zip file so that it will be tagged as updated whenever the update of data function is invoked.

```{r}
data_confirmed <- read_csv("data/time_series_covid19_confirmed_global.csv")
data_deceased  <- read_csv("data/time_series_covid19_deaths_global.csv")
current_date <- as.Date(names(data_confirmed)[ncol(data_confirmed)], format = "%m/%d/%y")
changed_date <- file_info("data/covid19_data.zip")$change_time

```

After that, to get the evolution data by country, we created another variable called **data_confirmed_sub** that chains **data_confirmed** with three functions which are **pivot_longer()**, **group_by()**, and **summarise()** in ordee to extract the relevant data from **data_confirmed**. The same measure is done on **data_deceased_sub** too in order to obtain data from **data_deceased**.

```{r}
data_confirmed_sub <- data_confirmed %>%
  pivot_longer(names_to = "date", cols = 5:ncol(data_confirmed)) %>%
  group_by(`Province/State`, `Country/Region`, date, Lat, Long) %>%
  summarise("confirmed" = sum(value, na.rm = T))

data_deceased_sub <- data_deceased %>%
  pivot_longer(names_to = "date", cols = 5:ncol(data_deceased)) %>%
  group_by(`Province/State`, `Country/Region`, date, Lat, Long) %>%
  summarise("deceased" = sum(value, na.rm = T))
```

After that, we created a variable named **data_evolution** which chains **data_confirmed_sub** with multiple functions such as **full_join()**, **ungroup()**, **mutate()**, **arrange()**, **group_by()**, **pivot_longer()** in order to get the evolution of cases based on time-series by country.

```{r}
data_evolution <- data_confirmed_sub %>%
  full_join(data_deceased_sub) %>%
  ungroup() %>%
  mutate(date = as.Date(date, "%m/%d/%y")) %>%
  arrange(date) %>%
  group_by(`Province/State`, `Country/Region`, Lat, Long) %>%
  mutate(
    recovered = lag(confirmed, 14, default = 0) - deceased,
    recovered = ifelse(recovered > 0, recovered, 0),
    active = confirmed - recovered - deceased
  ) %>%
  pivot_longer(names_to = "var", cols = c(confirmed, recovered, deceased, active)) %>%
  ungroup()
```

We also update the variable **data_evolution** by chaining it with **group_by()** and **mutate()** then **ungroup()** so that it will calculate new cases.

```{r}
data_evolution <- data_evolution %>%
  group_by(`Province/State`, `Country/Region`) %>%
  mutate(value_new = value - lag(value, 4, default = 0)) %>%
  ungroup()
```

After that, we remove all the data variables per invoke to clean up memory and to avoid duplicates when the next invocation occurs.

```{r}
rm(data_confirmed, data_confirmed_sub, data_deceased, data_deceased_sub)
```

We also downloaded the population data by using the **wb()** function which downloads data from the World Bank. We created a variable named **population** and grabbed the data from World Bank and assigned it to the variable. We chained the variable with two functions which are **select()** and **rename()**. After that, we created a variable named **countryNamesPop** which the columns consists of country names. We also created a variable named **countryNamesDat** which containes the official names of the countries.

```{r}
population <- wb(country = "countries_only", indicator = "SP.POP.TOTL", startdate = 2018, enddate = 2020) %>%
  select(country, value) %>%
  rename(population = value)
countryNamesPop <- c("Brunei Darussalam", "Congo, Dem. Rep.", "Congo, Rep.", "Czech Republic", "Egypt, Arab Rep.", "Iran, Islamic Rep.", "Korea, Rep.", "St. Lucia", "West Bank and Gaza", "Russian Federation", "Slovak Republic", "United States", "St. Vincent and the Grenadines", "Venezuela, RB")
countryNamesDat <- c("Brunei", "Congo (Kinshasa)", "Congo (Brazzaville)", "Czechia", "Egypt", "Iran", "Korea, South", "Saint Lucia", "occupied Palestinian territory", "Russia", "Slovakia", "US", "Saint Vincent and the Grenadines", "Venezuela")
population[which(population$country %in% countryNamesPop), "country"] <- countryNamesDat

```

For data that are not in John Hopkins University or the World Bank, we decided to scrap the data from Wikipedia. We created a simple data frame that consists of country and it's respective population numbers that is referenced from Wikipedia. We did that by creating a variable called **noDataCountries** and it will be a data frame. We then bind the **noDataCountries** with the original **population** by utilizing the **bind_rows()** function.

```{r}
noDataCountries <- data.frame(
  country    = c("Cruise Ship", "Guadeloupe", "Guernsey", "Holy See", "Jersey", "Martinique", "Reunion", "Taiwan*"),
  population = c(3700, 395700, 63026, 800, 106800, 376480, 859959, 23780452)
)
population      <- bind_rows(population, noDataCountries)

```

We then update the variable **data_evolution** by chaining **data_evolution** with a **left_join()** function. After that, we created a function named **date_atDate** which takes in the parameter of **inputDate**. In this function, we chain the selected element in **data_evolution** when the **data_evolution** object's date is equal to the **inputDate** which is the parameter of the function. We then chain this function with a **distinct()** function to keep distinct rows. We then continue the chain with **pivot_wider()** function that will expand the number of columns in **data_atData**. Last but not least, we finalized the function by linking the chain with **filter()** function that will select out rows with confirmed, recovered, deceased and active more than 0. Furthermore, we create a variable named **data_latest** which is the maximum evolution of the cases at the selected date.

```{r}
data_evolution <- data_evolution %>%
  left_join(population, by = c("Country/Region" = "country"))
rm(population, countryNamesPop, countryNamesDat, noDataCountries)

data_atDate <- function(inputDate) {
  data_evolution[which(data_evolution$date == inputDate),] %>%
    distinct() %>%
    pivot_wider(id_cols = c("Province/State", "Country/Region", "date", "Lat", "Long", "population"), names_from = var, values_from = value) %>%
    filter(confirmed > 0 |
             recovered > 0 |
             deceased > 0 |
             active > 0)
}

data_latest <- data_atDate(max(data_evolution$date))
```

We created a variable called **top5_countries** that consists of a chain of functions and variables which are **data_evolution**, **filter()**, **group_by()**, **summarise()**, **arrange()**, **top_n()**, **select()** and **pull()** in order to obtain the top 5 countries with the most active cases up until the day the application is launched.

```{r}
top5_countries <- data_evolution %>%
  filter(var == "active", date == current_date) %>%
  group_by(`Country/Region`) %>%
  summarise(value = sum(value, na.rm = T)) %>%
  arrange(desc(value)) %>%
  top_n(5) %>%
  select(`Country/Region`) %>%
  pull()
```

### Data Visualization

We first created the UI class that takes in each components separately to be shown on the main page. In this UI class, we inserted a navbar from the bootstrap class that contains to nav items which are **Overview** and **About**. We used CSS to style the components.

```{r}
source("UI/ui_overview.R", local = TRUE)
source("UI/ui_about.R", local = TRUE)


ui <- fluidPage(
  title = "COVID-19 Dashboard",
  tags$style(type = "text/css", ".container-fluid {padding-left: 0px; padding-right: 0px !important;}"),
  tags$style(type = "text/css", ".navbar {margin-bottom: 0px; background-color: #8B0000}"),
  tags$style(type = "text/css", ".content {padding: 0px;}"),
  tags$style(type = "text/css", ".row {margin-left: 0px; margin-right: 0px;}"),
  tags$style(HTML(".col-sm-12 { padding: 5px; margin-bottom: -15px; }")),
  tags$style(HTML(".col-sm-6 { padding: 5px; margin-bottom: -15px; }")),
  navbarPage(
    title       = div("COVID-19 Dashboard", style = "padding-left: 10px"),
    inverse = TRUE,
    collapsible = TRUE,
    fluid       = TRUE,
    tabPanel("Overview", page_overview, value = "page-overview"),
    tabPanel("About", page_about, value = "page-about"),
    tags$script(HTML("var header = $('.navbar > .container-fluid');
    console.log(header)")
    )
  )
)
```

For the visualization part, we have 4 value boxes which consists of **Confirmed**, **Estimated Recoveries**, **Deceased**, and **Affected Countries** respectively. The valuebox were coded in **keyFigures.R**. We first created a function called **sumData** that takes in a parameter called **date**. The function basically returns a summary of the confirmed, recovered, deceased, and countries affected data if the current date is greater or equal than the **data_evoution** date. After the data was summarised, the function returns the data after exiting the condition.

```{r}
sumData <- function(date) {
  if (date >= min(data_evolution$date)) {
    data <- data_atDate(date) %>% summarise(
      confirmed = sum(confirmed, na.rm = T),
      recovered = sum(recovered, na.rm = T),
      deceased  = sum(deceased, na.rm = T),
      countries = n_distinct(`Country/Region`)
    )
    return(data)
  }
  return(NULL)
}
```

After that, we created a variable called **key_figures** that consists of a **reactive()** function that takes input from the widget slider and invokes the previously created function which is **sumData()**. The variables **data** and **data_yesterday** were then updated. Then we created a variable called **data_new** that takes in a list of variables called **new_confirmed**, **new_recovered** and **new_deceased** that takes the current data subtract the yesterday's data over the data from yesterday multiplied with 100. We then created another variable called **keyFigures** which accommodates a list of HTML formatted headers extracting data from the previous **data_new** list.

```{r}

key_figures <- reactive({
  data           <- sumData(input$timeSlider)
  data_yesterday <- sumData(input$timeSlider - 1)

  data_new <- list(
    new_confirmed = (data$confirmed - data_yesterday$confirmed) / data_yesterday$confirmed * 100,
    new_recovered = (data$recovered - data_yesterday$recovered) / data_yesterday$recovered * 100,
    new_deceased  = (data$deceased - data_yesterday$deceased) / data_yesterday$deceased * 100,
    new_countries = data$countries - data_yesterday$countries
  )

  keyFigures <- list(
    "confirmed" = HTML(paste(format(data$confirmed, big.mark = " "), sprintf("<h4>(%+.1f %%)</h4>", data_new$new_confirmed))),
    "recovered" = HTML(paste(format(data$recovered, big.mark = " "), sprintf("<h4>(%+.1f %%)</h4>", data_new$new_recovered))),
    "deceased"  = HTML(paste(format(data$deceased, big.mark = " "), sprintf("<h4>(%+.1f %%)</h4>", data_new$new_deceased))),
    "countries" = HTML(paste(format(data$countries, big.mark = " "), "/ 195", sprintf("<h4>(%+d)</h4>", data_new$new_countries)))
  )
  return(keyFigures)
})


```

After that, we output the value boxes for every category from **Confirmed** to **Affected Countries**, we gave it different colors and linked it to differenct attributes of **key_figures()**. They all had a bootstrap width of 3 so that their total width will be 12 which is the max of bootstrap width. 

```{r}
output$valueBox_confirmed <- renderValueBox({
  valueBox(
    key_figures()$confirmed,
    subtitle = "Confirmed",
    icon     = icon("ambulance"),
    color    = "red",
    width    = NULL
  )
})


output$valueBox_recovered <- renderValueBox({
  valueBox(
    key_figures()$recovered,
    subtitle = "Estimated Recoveries",
    icon     = icon("file-medical"),
    color    = "green"
  )
})

output$valueBox_deceased <- renderValueBox({
  valueBox(
    key_figures()$deceased,
    subtitle = "Deceased",
    icon     = icon("skull-crossbones"),
    color    = "purple"
  )
})

output$valueBox_countries <- renderValueBox({
  valueBox(
    key_figures()$countries,
    subtitle = "Affected Countries",
    icon     = icon("flag"),
    color    = "blue"
  )
})

output$box_keyFigures <- renderUI(box(
  title = paste0("Key Figures (", strftime(input$timeSlider, format = "%d.%m.%Y"), ")"),
  fluidRow(
    column(
      valueBoxOutput("valueBox_confirmed", width = 3),
      valueBoxOutput("valueBox_recovered", width = 3),
      valueBoxOutput("valueBox_deceased", width = 3),
      valueBoxOutput("valueBox_countries", width = 3),
      width = 12,
      style = "margin-left: -20px"
    )
  ),
  div("Last updated: ", strftime(changed_date, format = "%d.%m.%Y - %R %Z")),
  width = 12
))
```

For the summary table, we render a tab box and a table. The figures of the summary change dynamically with the slider widget due to the implementation of time slider.

```{r}
output$summaryTables <- renderUI({
  tabBox(
    tabPanel("Country/Region",
      div(
        dataTableOutput("summaryDT_country"),
        style = "margin-top: -10px")
    ),
    width = 12
  )
})

output$summaryDT_country <- renderDataTable(getSummaryDT(data_atDate(current_date), "Country/Region", selectable = TRUE))
proxy_summaryDT_country  <- dataTableProxy("summaryDT_country")

observeEvent(input$timeSlider, {
  data <- data_atDate(input$timeSlider)
  replaceData(proxy_summaryDT_country, summariseData(data, "Country/Region"), rownames = FALSE)
}, ignoreInit = TRUE, ignoreNULL = TRUE)

observeEvent(input$summaryDT_country_row_last_clicked, {
  selectedRow     <- input$summaryDT_country_row_last_clicked
  selectedCountry <- summariseData(data_atDate(input$timeSlider), "Country/Region")[selectedRow, "Country/Region"]
  location        <- data_evolution %>%
    distinct(`Country/Region`, Lat, Long) %>%
    filter(`Country/Region` == selectedCountry) %>%
    summarise(
      Lat  = mean(Lat),
      Long = mean(Long)
    )
  leafletProxy("overview_map") %>%
    setView(lng = location$Long, lat = location$Lat, zoom = 4)
})

summariseData <- function(df, groupBy) {
  df %>%
    group_by(!!sym(groupBy)) %>%
    summarise(
      "Confirmed"            = sum(confirmed, na.rm = T),
      "Estimated Recoveries" = sum(recovered, na.rm = T),
      "Deceased"             = sum(deceased, na.rm = T),
      "Active"               = sum(active, na.rm = T)
    ) %>%
    as.data.frame()
}

getSummaryDT <- function(data, groupBy, selectable = FALSE) {
  datatable(
    na.omit(summariseData(data, groupBy)),
    rownames  = FALSE,
    options   = list(
      order          = list(1, "desc"),
      scrollX        = TRUE,
      scrollY        = "37vh",
      scrollCollapse = T,
      dom            = 'ft',
      paging         = FALSE
    ),
    selection = ifelse(selectable, "single", "none")
  )
}
```

For the map label, we first created the **addLabel()** function that takes **data** as parameter. It will appear as a tooltip when the pointer is hovered over the marker, displaying essential info such as **confirmed**, **deceased**, **recovered**, and **active**.

```{r}
library("htmltools")

addLabel <- function(data) {
  data$label <- paste0(
    '<b>', ifelse(is.na(data$`Province/State`), data$`Country/Region`, data$`Province/State`), '</b><br>
    <table style="width:120px;">
    <tr><td>Confirmed:</td><td align="right">', data$confirmed, '</td></tr>
    <tr><td>Deceased:</td><td align="right">', data$deceased, '</td></tr>
    <tr><td>Estimated Recoveries:</td><td align="right">', data$recovered, '</td></tr>
    <tr><td>Active:</td><td align="right">', data$active, '</td></tr>
    </table>'
  )
  data$label <- lapply(data$label, HTML)

  return(data)
}

```

After that, we created a variable called **map** and chain a series of functions onto it. We added layer controls to visualize different types of data such as **confirmed**, **confirmed per capita**, **estimated recoveries**, **deceased**, **active**, **active per capita**. We also added a custom input time slider to show the data dynamically through the time series. To visualize different types of data output, we used blue for confirmed cases, dark blue for confirmed cases per capita,green for estimated recoveries, black for deceased, red for active and dark red for active per capita. Finally, we render the map.

```{r}
map <- leaflet(addLabel(data_latest)) %>%
  setMaxBounds(-180, -90, 180, 90) %>%
  setView(0, 20, zoom = 2) %>%
  addTiles() %>%
  addProviderTiles(providers$CartoDB.Positron, group = "Light") %>%
  addProviderTiles(providers$HERE.satelliteDay, group = "Satellite") %>%
  addLayersControl(
    baseGroups    = c("Satellit", "Light"),
    overlayGroups = c("Confirmed", "Confirmed (per capita)", "Estimated Recoveries", "Deceased", "Active", "Active (per capita)")
  ) %>%
  hideGroup("Confirmed (per capita)") %>%
  hideGroup("Estimated Recoveries") %>%
  hideGroup("Deceased") %>%
  hideGroup("Active") %>%
  hideGroup("Active (per capita)") %>%
  addEasyButton(easyButton(
    icon    = "glyphicon glyphicon-globe", title = "Reset zoom",
    onClick = JS("function(btn, map){ map.setView([20, 0], 2); }"))) %>%
  addEasyButton(easyButton(
    icon    = "glyphicon glyphicon-map-marker", title = "Locate Me",
    onClick = JS("function(btn, map){ map.locate({setView: true, maxZoom: 6}); }")))

observe({
  req(input$timeSlider, input$overview_map_zoom)
  zoomLevel               <- input$overview_map_zoom
  data                    <- data_atDate(input$timeSlider) %>% addLabel()
  data$confirmedPerCapita <- data$confirmed / data$population * 100000
  data$activePerCapita    <- data$active / data$population * 100000

  leafletProxy("overview_map", data = data) %>%
    clearMarkers() %>%
    addCircleMarkers(
      lng          = ~Long,
      lat          = ~Lat,
      radius       = ~log(confirmed^(zoomLevel / 2)),
      stroke       = FALSE,
      color        = "#add8e6",
      fillOpacity  = 0.5,
      label        = ~label,
      labelOptions = labelOptions(textsize = 15),
      group        = "Confirmed"
    ) %>%
    addCircleMarkers(
      lng          = ~Long,
      lat          = ~Lat,
      radius       = ~log(confirmedPerCapita^(zoomLevel)),
      stroke       = FALSE,
      color        = "#0000FF",
      fillOpacity  = 0.5,
      label        = ~label,
      labelOptions = labelOptions(textsize = 15),
      group        = "Confirmed (per capita)"
    ) %>%
    addCircleMarkers(
      lng          = ~Long,
      lat          = ~Lat,
      radius       = ~log(recovered^(zoomLevel)),
      stroke       = FALSE,
      color        = "#00FF00",
      fillOpacity  = 0.5,
      label        = ~label,
      labelOptions = labelOptions(textsize = 15),
      group = "Estimated Recoveries"
    ) %>%
    addCircleMarkers(
      lng          = ~Long,
      lat          = ~Lat,
      radius       = ~log(deceased^(zoomLevel)),
      stroke       = FALSE,
      color        = "#000000",
      fillOpacity  = 0.5,
      label        = ~label,
      labelOptions = labelOptions(textsize = 15),
      group        = "Deceased"
    ) %>%
    addCircleMarkers(
      lng          = ~Long,
      lat          = ~Lat,
      radius       = ~log(active^(zoomLevel / 2)),
      stroke       = FALSE,
      color        = "#FF0000",
      fillOpacity  = 0.5,
      label        = ~label,
      labelOptions = labelOptions(textsize = 15),
      group        = "Active"
    ) %>%
    addCircleMarkers(
      lng          = ~Long,
      lat          = ~Lat,
      radius       = ~log(activePerCapita^(zoomLevel)),
      stroke       = FALSE,
      color        = "#8b0000",
      fillOpacity  = 0.5,
      label        = ~label,
      labelOptions = labelOptions(textsize = 15),
      group        = "Active (per capita)"
    )
})

output$overview_map <- renderLeaflet(map)
```

### Conclusion

As we can see, COVID-19 has no means of stopping and the trend is still increasing worldwide with it having no intentions of stopping its massacre. Hence, we must break the chain of infection through vaccination, getting the latest info and prevent outings or gatherings at public places. TOgether, we can bring down COVID-19 and restore our freedom.

